\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

This report presents \textbf{Cognitive Copilot}, a local-first, full-stack academic Learning Management System that integrates state-of-the-art AI research directly into the course-study workflow. The system is designed for university deployment without cloud LLM subscriptions, running entirely on commodity hardware using the Ollama runtime with \texttt{qwen2.5:7b-instruct} for generation and \texttt{nomic-embed-text} for dense retrieval.

The platform's technical core comprises five interlocking contributions. First, a \textbf{production-grade Retrieval-Augmented Generation (RAG)} pipeline stores 768-dimensional embeddings in \texttt{pgvector} and lexical BM25 indexes in Postgres \texttt{tsvector}, fusing both ranking signals via Reciprocal Rank Fusion before streaming a cited answer to the student. Second, a \textbf{dual-mode OCR system} keeps Tesseract as the fast default while delegating to a Python FastAPI sidecar — TrOCR for handwriting, Nougat for academic PDFs with embedded mathematics — via a dedicated BullMQ worker so slow inferences never block the HTTP thread. Third, a \textbf{Socratic agentic tutor} implements five pedagogical strategies (Explain, Socratic, Hint Ladder, Worked Example, Misconception Probe) selected by a deterministic rule system, with generation carried out through a bounded ReAct loop (five iterations, twelve seconds wall-clock) equipped with five callable tools. Fourth, \textbf{Bayesian Beta-posterior mastery tracking} replaces a naive heuristic by maintaining conjugate $\mathrm{Beta}(\alpha,\beta)$ hyperparameters per (student, topic) pair; the posterior mean and a 95\% credible interval are surfaced in the quiz UI so students can read their own uncertainty. Fifth, an \textbf{LLM observability and evaluation harness} logs every Ollama call to a structured \texttt{LlmCall} table, publishes Prometheus metrics, and runs an automated golden-dataset evaluation reporting recall@5, faithfulness, and OCR character-error rate.

The backend is implemented with Express and TypeScript backed by PostgreSQL~15 and Redis~7; the frontend uses React~19 with Vite, TanStack Query, Zustand, and a bespoke shadcn-style design system with dark mode. Full production packaging is provided through multi-stage Dockerfiles and Compose profiles. The evaluation harness reports recall@5~=~0.83, faithfulness~=~0.79 (LLM-as-judge, majority of three), and a handwriting OCR word error rate of 0.27 on the accurate path versus 0.83 on the fast path.

The project demonstrates that a capstone engineering team can build a measurably better tutoring experience by applying established ML research (RAG, BKT, ReAct) within a disciplined software-engineering process, without relying on paid cloud APIs.
