\chapter{Introduction}

\section{Background}
University students routinely juggle fragmented digital tools: a separate platform for course materials, another for the class schedule, a third for discussion, and a generic chatbot for study help. None of these tools know about one another, so continuity of learning is lost whenever a student switches context. Cognitive Copilot was designed to collapse this fragmentation into a single, coherent web application that understands courses, materials, and the student's measured mastery level simultaneously.

The system is also designed as a deliberate counterpoint to cloud-LLM-dependent tutors. By running all inference locally through the Ollama runtime, it is deployable on a university server or even a faculty laptop without recurring API subscription costs. Local-first operation also removes the data-privacy concerns that arise when student conversations are sent to external APIs.

\section{Problem Statement}
Three concrete gaps motivate this work.

\begin{enumerate}[label=\arabic*.]
\item \textbf{Grounding gap.} General-purpose LLM chat has no knowledge of a specific course's materials, so it cannot cite a passage from a lecture PDF or flag that an answer contradicts the textbook.
\item \textbf{Pedagogy gap.} Most LLM wrappers use a single system prompt for all interactions. There is no strategy selection, no Socratic probing, and no distinction between a student who is confused and one who is ready for a worked example.
\item \textbf{Measurability gap.} Without instrumentation, a supervisor cannot tell whether a tutoring system is helping students or generating confident nonsense. There is no retrieval recall, no faithfulness score, and no per-student mastery trajectory.
\end{enumerate}

Cognitive Copilot addresses all three gaps within a single deployable system.

\section{Objectives}
\begin{enumerate}[label=\arabic*.]
\item Build a secure, role-aware full-stack LMS covering authentication, course-topic-material management, community discussion, and exam analytics.
\item Implement a \textbf{production RAG pipeline} using \texttt{pgvector} for 768-dim dense retrieval and Postgres \texttt{tsvector} for BM25 sparse retrieval, fused via Reciprocal Rank Fusion, with streaming cited answers.
\item Build a \textbf{Socratic agentic tutor} with five pedagogical strategies selected by a deterministic rule system, executed through a bounded ReAct loop with five callable tools.
\item Replace the existing na\"{i}ve mastery heuristic with \textbf{Bayesian Beta-posterior tracking} per (student, topic), surfacing a 95\% credible interval in the quiz UI.
\item Implement a \textbf{dual-mode OCR pipeline}: a fast Tesseract path for routine digitization and an accurate Python-sidecar path (TrOCR, Nougat) for handwriting and academic PDFs.
\item Build an \textbf{LLM observability and evaluation harness}: per-call \texttt{LlmCall} audit table, Prometheus metrics endpoint, and an automated golden-dataset evaluator reporting recall@K, faithfulness, and OCR WER.
\item Deliver \textbf{production packaging}: multi-stage Dockerfiles, Compose profiles for development and demo, and a GitHub Actions CI pipeline with lint, typecheck, tests, and eval smoke-test.
\end{enumerate}

\section{Significance of the Project}
The significance lies in the combination. RAG, Socratic tutoring, Bayesian knowledge tracing, and agentic tool use are individually well-studied; what is unusual is implementing all four in a single, locally-deployable system with an automated evaluation harness that lets a supervisor verify the claims at viva time. The engineering approach --- modular route-controller-service layers, worker-based async pipelines, and structured LLM outputs --- also serves as a reference design for future capstone projects building on AI infrastructure.

\section{Project Scope}
The implemented system covers: user authentication and profiles; course, topic, material, and enrollment management; AI tutor with five modes (Chat, Ask Course, Explain, Quiz, Resource Discovery); routine scan via OCR; community discussion and announcements; attendance and marks tracking; real-time notifications via Socket.IO; analytics and the instructor evaluation dashboard; and a complete Docker deployment stack. Out of scope: mobile native clients, Bengali handwriting OCR, and multi-tenant isolation.

\section{Novelty of the Work}
The primary novelty is the integration of (a) grounded retrieval-augmented generation, (b) advanced handwriting and academic-PDF OCR, (c) a Socratic agent with deterministic strategy selection, (d) Bayesian mastery tracking, and (e) an automated evaluation harness --- all running offline without cloud LLM APIs. No open-source capstone project we are aware of combines all five in a single deployable codebase.

\section{Project Planning}
Implementation was carried out over four iterative phases corresponding to the plan outlined in the project proposal: infrastructure and RAG foundations, agent tutor and OCR sidecar, evaluation and design system, and user study and report. Figure~\ref{fig:gantt} presents the activity timeline.

\begin{figure}[H]
\centering
\begin{tikzpicture}[x=1.05cm, y=0.82cm]
  \draw[->] (0,0) -- (12.0,0) node[right] {Weeks};
  \foreach \x in {1,...,11} {
    \draw (\x,0.08) -- (\x,-0.08);
    \node[below,font=\scriptsize] at (\x,-0.10) {\x};
  }
  \node[left,align=right,font=\small] at (0,2.1) {Requirements \&\\Architecture};
  \node[left,align=right,font=\small] at (0,1.3) {RAG + Observ.};
  \node[left,align=right,font=\small] at (0,0.6) {Agent + OCR};
  \node[left,align=right,font=\small] at (0,-0.1) {Eval + Report};

  \fill[blue!40]   (0.8,1.95) rectangle (2.8,2.25);
  \fill[teal!45]   (2.5,1.15) rectangle (5.0,1.45);
  \fill[orange!50] (4.6,0.45) rectangle (8.2,0.75);
  \fill[purple!40] (7.8,-0.25) rectangle (11.0,0.05);
\end{tikzpicture}
\caption{Project Activity Timeline (four iterative phases over eleven weeks)}
\label{fig:gantt}
\end{figure}

\section{Work Distribution}
\begin{itemize}
\item \textbf{Khadimul Islam Mahi (2107076):} Backend architecture, RAG pipeline (chunker, embedder, retriever), AI agent (strategy selector, ReAct loop, tools), OCR sidecar integration, LLM observability, BullMQ workers, evaluation harness, Docker and CI pipeline, LaTeX report.
\item \textbf{Sumaiya Akter (2107080):} Frontend pages (AI tutor modes, analytics, landing, evaluation dashboard), design system primitives, dark mode, onboarding tour, TanStack Query data layer, Socket.IO notification hooks, UI integration testing.
\item \textbf{Joint:} Database schema design, seed data, user study protocol, bug-fixing, and final system validation.
\end{itemize}

\section{Applications}
The system is applicable to university department-level deployments where faculty upload course materials and students interact with an AI tutor grounded in those materials. The OCR routine scanner addresses a concrete Bangladesh university need: paper timetables photographed and digitized into a structured weekly schedule. The evaluation dashboard gives supervisors an auditable view of tutoring quality without relying on anecdotes.

\section{Report Organization}
Chapter~II reviews related works across RAG, document OCR, intelligent tutoring, and agent architectures. Chapter~III presents the full methodology including system architecture, RAG pipeline, OCR sidecar, ReAct agent, and Bayesian mastery model. Chapter~IV describes implementation, results, and the quantitative evaluation. Chapter~V discusses societal and ethical considerations. Chapter~VI maps the work to complex engineering problem criteria. Chapter~VII concludes with limitations and future work.
