\chapter{Related Works}

The Cognitive Copilot project sits at the intersection of four distinct research communities: retrieval-augmented generation (RAG), document and handwriting OCR, intelligent tutoring systems (ITS), and agentic large-language-model (LLM) design. This chapter situates our system against each of these, identifies concrete gaps, and shows how the design decisions in Chapter~\ref{ch:design} follow directly from that positioning.

\section{Digital Learning Management Systems}
General-purpose LMS platforms such as Moodle, Google Classroom, and Canvas remain the dominant choice for university course management. These systems excel at assignment flow, enrolment, and grading, but their AI capability is either absent or bolted on through premium cloud add-ons \cite{react,express,postgres}. Few offer local tutor integration, and none expose the retrieval primitives needed to ground an answer in specific course material. Cognitive Copilot deliberately narrows the feature surface compared to a full LMS and instead invests the engineering budget in pedagogy, grounding, and observability.

\section{Retrieval-Augmented Generation for Education}
Lewis et al.~\cite{lewis2020rag} introduced RAG as a way to fuse a non-parametric retrieval module with a parametric sequence-to-sequence generator, showing that generation quality on knowledge-intensive tasks improves substantially when the model can attend over a retrieved passage set. The survey by Gao et al.~\cite{gao2023ragsurvey} consolidates the current design space: chunk-level retrieval, hybrid sparse/dense scoring, reranking, and increasingly compositional retrieval pipelines.

For the education domain in particular, RAG enables two properties that free-form LLMs cannot provide: (i)~answers traceable to a cited source, and (ii)~reduced hallucination because the generator is constrained by the retrieved evidence. Cognitive Copilot builds on four specific pieces of this literature:

\begin{enumerate}[label=\arabic*.]
    \item \textbf{Dense retrieval} via dual-encoder embeddings \cite{karpukhin2020dpr}. We use \texttt{nomic-embed-text} \cite{nomic}, a 768-dimensional open-weights model chosen for its strong MTEB scores at a CPU-friendly size.
    \item \textbf{Sparse retrieval} via BM25 \cite{robertson2009bm25}, implemented over Postgres \texttt{tsvector} columns so we avoid introducing a separate search service.
    \item \textbf{Hybrid fusion} via Reciprocal Rank Fusion \cite{cormack2009rrf}, which combines dense and sparse rankings without requiring score calibration.
    \item \textbf{Vector storage} via \texttt{pgvector} \cite{pgvector} with HNSW indexing \cite{malkov2018hnsw}, keeping our persistence stack to a single database rather than a dedicated vector store.
\end{enumerate}

\section{Handwriting and Document OCR}
Tesseract remains a capable engine for printed text \cite{tesseract}, but performs poorly on handwritten and mathematical content. Two recent transformer-based models substantially change the landscape. Li et al.~\cite{li2023trocr} introduced TrOCR, which combines an image transformer encoder with a text decoder and reaches production-quality recognition on handwritten text in Latin scripts. Blecher et al.~\cite{blecher2023nougat} introduced Nougat, which extends this idea to scientific PDFs, producing markdown with embedded LaTeX directly from raster page images. The Unstructured toolkit \cite{unstructured} complements both by providing layout-aware segmentation for complex PDFs.

Our system exposes the fast Tesseract path as the default and delegates to a Python FastAPI sidecar for TrOCR and Nougat when the instructor selects the ``accurate'' quality option on upload. The sidecar isolates the heavy Python dependency tree from the Node.js backend and lets us version those models independently. A known limitation reported openly in Chapter~\ref{ch:limitations}: TrOCR's pretrained checkpoint is Latin-only, so Bengali handwriting remains unsupported.

\section{Intelligent Tutoring Systems and Mastery Modelling}
Classical ITS research \cite{anderson1995cognitive,vanlehn2011tutors} showed that structured tutoring with explicit mastery tracking outperforms unstructured help in controlled studies. Corbett and Anderson~\cite{corbett1994bkt} formalised mastery as a latent binary variable updated through Bayesian inference in the influential Bayesian Knowledge Tracing (BKT) model. Piech et al.~\cite{piech2015dkt} generalised this to sequence models in Deep Knowledge Tracing. Wilson et al.~\cite{wilson2016estimating} showed that simpler probabilistic models often match deep alternatives when per-student data is sparse --- the regime in which a small capstone user study operates.

Cognitive Copilot adopts a minimal but principled stance: a conjugate Beta posterior per (student, topic) pair, updated after every quiz attempt. The posterior mean drives difficulty targeting, and the 95\% credible interval is surfaced in the UI. This is deliberately simpler than BKT, but stronger than the na\"{i}ve $+0.1 / -0.05$ rule used in many capstone tutors --- a choice we justify quantitatively in Chapter~\ref{ch:evaluation}.

\section{Socratic Dialogue and Pedagogy}
Human tutors typically withhold a direct answer on first ask and instead probe for the student's mental model \cite{chi2001socratic}. Commercial tutors such as Khanmigo \cite{khanmigo2023} have popularised Socratic prompting, but few open-source systems implement it explicitly with multiple named strategies. We encode five pedagogical strategies --- Explain, Socratic, Hint Ladder, Worked Example, and Misconception Probe --- and select among them with a deterministic rule-based selector augmented by the student's current posterior mean. The full rule set is detailed in Section~\ref{sec:agent}.

\section{Agent Architectures and Tool Use}
Chat-only tutors cannot perform concrete actions such as searching the web or running code to verify a student's hypothesis. The ReAct paradigm \cite{yao2023react} showed that interleaving Thought and Action tokens in the same context enables LLMs to use external tools, and Toolformer \cite{schick2023toolformer} demonstrated that tool-call synthesis can be self-supervised. We implement a bounded ReAct loop (maximum five iterations, twelve seconds wall-clock) with five tools: \texttt{retrieve\_material}, \texttt{search\_web}, \texttt{generate\_practice\_question}, \texttt{render\_diagram}, and \texttt{run\_code}. Structured outputs are produced via Ollama's JSON-schema-constrained \texttt{format} parameter, removing the brittle regex parsing that plagued the earlier iteration of the quiz generator.

\section{Evaluating Generative Systems}
Evaluating RAG requires measuring both retrieval quality and answer quality jointly. RAGAS \cite{es2023ragas} and ARES \cite{saad2023arespolicy} popularised LLM-as-judge metrics for faithfulness, context precision, and answer relevance. We adopt a simplified in-repo variant: a three-times majority-vote LLM judge for faithfulness, keyword-overlap for answer relevance, and standard recall@K for retrieval quality. Complete definitions and the golden dataset are presented in Chapter~\ref{ch:evaluation}.

\section{Position of This Project}
Cognitive Copilot does not claim novelty in any one of the above areas individually; its contribution is \emph{integration}. No other open-source capstone we are aware of combines (a)~local-first RAG over student-uploaded materials, (b)~handwriting and academic-PDF OCR, (c)~a Socratic agent with tool use, (d)~Bayesian mastery tracking, and (e)~an end-to-end evaluation harness --- all running on a single developer machine without cloud LLM APIs. Table~\ref{tab:comparison} summarises the differentiator grid.

\begin{table}[H]
\centering
\small
\caption{Feature comparison against representative learning systems. $\checkmark$ = fully implemented, $\circ$ = partial/vendor-locked, -- = absent.}
\label{tab:comparison}
\begin{tabular}{lcccc}
\toprule
\textbf{Feature} & \textbf{Moodle} & \textbf{ChatGPT Edu} & \textbf{Khanmigo} & \textbf{Ours} \\
\midrule
Grounded retrieval + citations & --           & $\circ$     & $\checkmark$ & $\checkmark$ \\
Handwriting OCR                & --           & --          & --           & $\checkmark$ \\
Math OCR (LaTeX output)        & --           & --          & --           & $\checkmark$ \\
Socratic strategy selector     & --           & --          & $\checkmark$ & $\checkmark$ \\
Bayesian mastery + CI          & --           & --          & --           & $\checkmark$ \\
Tool-using agent               & --           & $\circ$     & --           & $\checkmark$ \\
100\% local / no cloud LLM     & $\checkmark$ & --          & --           & $\checkmark$ \\
Open evaluation harness        & --           & --          & --           & $\checkmark$ \\
Internet resource discovery    & --           & $\circ$     & --           & $\checkmark$ \\
Real-time notifications (WS)   & $\checkmark$ & --          & --           & $\checkmark$ \\
\bottomrule
\end{tabular}
\end{table}
